---
title: "607 Project 1"
author: "Randi Skrelja"
date: "September 27, 2015"
---
In this project, you're given a text file with chess tournament results where the information has some structure. Your
job is to create an R Markdown file that generates a .CSV file (that could for example be imported into a SQL database)

With the following information for all of the players:

Player's Name, Player's State, Total Number of Points, Player's Pre-Rating, and Average Pre Chess Rating of Opponents.

Note: For the first player, the information would be: Gary Hua, ON, 6.0, 1794, 1605; 1605 was calculated by using the
pre-tournament opponents' ratings of 1436, 1563, 1600, 1610, 1649, 1663, 1716, and dividing by the total number of games played.

```{r}
# Read data file
chess <- read.csv("https://raw.githubusercontent.com/vskrelja/607_DataAcqMgt_Skrelja/master/tournamentinfo.txt", skip=3)
```

```{r}
library(stringr)

# Player's Name
name <- str_trim(unlist(str_extract_all(unlist(chess), "\\w+(\\ \\w+ (\\w|-)* \\w*)")))

# Player's State
state <- str_trim(unlist(str_extract_all(unlist(chess), " ON | MI | OH ")))

# Total Number of Points
points <- str_trim(unlist(str_extract_all(unlist(chess), "\\d(\\.\\d)")))

# Player's Pre-rating
pre_rating <- str_replace_all(str_trim(unlist(str_extract_all(unlist(chess), "R: [[:digit:] ]*"))), "R: ", "")

# Average Pre Chess Rating of Opponents
x <- str_sub(unlist(chess), start = 48, end = 89)[c(seq(1, length(unlist(chess)), by = 3))]
opp_rating <- str_extract_all(x, "( |\\d){4}")
opp_rating <- as.numeric(unlist(opp_rating))
opp_rating <- as.numeric(pre_rating[opp_rating]) 
opp_rating_mtx <- matrix(opp_rating, nrow=7, ncol=64)
opp_rating_avg <- colMeans(opp_rating_mtx, na.rm=T)

# Data Frame
data.frame(name, state, points, pre_rating, opp_rating_avg)
```
